Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-02-19 paper/main.tex
Wrote section describing iterative alignment
computation, including a couple new algorithms.
Also added back in my discussion of computational
complexity.

While solving \eqref{eq:l1min_denoise} can achieve exceptional
recognition accuracy on public databases, its success relies on the assumption
that the query image is well aligned with the training images. In
\cite{WagnerA2009-CVPR}, the framework of SBC was extended to iteratively align
a query image to one subject class. The face alignment algorithm in turn solves
the following minimization problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain, which aligns $\bb$
with the training images from the $i$-th class $A_i$. The algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} to both to use the
$\ell_1$ norm as a robust error function, and also simultaneously 
to estimate the illumination model $A\x$.  
The original solution \cite{WagnerA2009-CVPR} was implemented based 
on standard interior-point
methods, and its speed has been since improved by ALM in a more recent paper
\cite{WagnerA2011-PAMI}. However, primarily due the high per-user cost
of the alignment step, the recognition system was still not fast enough
for real-time recognition against datasets of hundreds of users.
In this paper, in addition to solving the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
discuss how to properly accelerate the iterative face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

@@

\section{Parallelization of the Face Alignment Stage}
\label{sec:alignment}
{\bf AY: Add an algorithm here describe the alignment block in the pipeline. Explain the algorithm and necessary modification from generic l-1}

>>>>

While solving \eqref{eq:l1min_denoise} can achieve exceptional
recognition accuracy on public databases, its success relies on the assumption
that the query image is well aligned with the training images. In
\cite{WagnerA2009-CVPR}, the framework of SBC was extended to iteratively align
a query image to one subject class. The face alignment algorithm in turn solves
the following minimization problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \tilde\bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain, which aligns $\tilde\bb$
with the training images from the $i$-th class $A_i$. The algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} to both to use the
$\ell_1$ norm as a robust error function, and also simultaneously 
to estimate the illumination model $A\x$.  
The original solution \cite{WagnerA2009-CVPR} was implemented based 
on standard interior-point
methods, and its speed has been since improved by ALM in a more recent paper
\cite{WagnerA2011-PAMI}. However, primarily due the high per-user cost
of the alignment step, the recognition system was still not fast enough
for real-time recognition against datasets of hundreds of users.
In this paper, in addition to solving the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
discuss how to properly accelerate the iterative face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

@@

\section{Parallelization of the Face Alignment Stage}
\label{sec:alignment}
%{\bf AY: Add an algorithm here describe the alignment block in the pipeline. Explain the algorithm and necessary modification from generic l-1}
This section provides a brief explanation of the iterative alignment stage
of the recognition pipeline.  
The inputs of the algorithm are the test image $f(x,y)$ stacked into an $M$-
dimensional vector $\f$, and the aligned and stacked training images $A_i$ for
each training user $i$. 
The algorithm repeatedly solves a linearized version of the minimization problem
in \eqref{eq:l1min_alignment}.  This involves two computational steps:
the computation of the linearization, and the solution of the minimization problem.

The computation of the linearization is complicated by the need to avoid trivial 
solutions where the algorithm
progressively zooms in on a black pixel of the test image.
This is achieved by minimizing the $\ell_1$ error between the illumination model $Ax$ and the cropped,
resampled, and normalized test image $\tilde\bb$, rather than the un-normalized version as in 
conventional Lucas-Kanade iterative alignment.
This involves a bit of calculus:
\begin{equation}
\frac{\partial}{\partial \tau_p} \frac{\bb}{||\bb||} 
= \frac{1}{||\bb||} \frac{\partial \bb}{\partial \tau_p} - 
\frac{\bb}{||\bb||^3} \bb^T \frac{\partial \bb}{\partial \tau_p}
\end{equation}
If the warping $\tau$ is constrained to be a similarity transform,
it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
once per test image. {bf DW: I could go into the rest of the math for this, but I'm worried about space}

The iterative alignment stage is summarized as Algorithm \ref{alg:iterative_alignment}.
\begin{algorithm}[thb]
\caption{\bf (Iterative Alignment Stage)} \label{alg:iterative_alignment}
\begin{algorithmic}[1]
\begin{small}
\STATE {\bf Input:} Aligned training images $\{A_i \in \Re^{m\times n_i}\}_{i=1}^K$ for $K$ subjects,  a test image
$\f\in\Re^M$ 
\STATE $\tau^{(0)} \leftarrow \textrm{face detector($\f$)}$. 
\STATE $\f_x, \f_y \in \Re^M \leftarrow \frac{df}{dx}, \frac{df}{dy}$ 
\STATE {\bf for} each gallery subject $i$, 
\STATE \hspace{3mm} {\bf while} not converged $(j=1,2,\ldots)$ {\bf do} 
\STATE \hspace{6mm} $\tilde \bb(\tau) \leftarrow \frac{\bb \circ \tau}{\|\bb \circ \tau\|_2}$;
\STATE \hspace{6mm} $J \leftarrow  \frac{\partial}{\partial \tau} \tilde\bb(\tau)  \bigr|_{\tau^{(j)}} $;
\STATE \hspace{6mm} $\Delta \tau =  \arg\min \; \| \e \|_1  \; \subj \; \tilde \bb + J \Delta \tau = A_i \x + \e.$ 
\STATE \hspace{6mm} $\tau^{(j+1)} \leftarrow \tau^{(j)} + \Delta
\tau$; 
\end{small}
\end{algorithmic}
\end{algorithm}

The solution of the minimization problem in Algorithm
\ref{alg:iterative_alignment} Step 8 can be solved via a
variant of ALM, which is summarized as Algorithm \ref{alg:alignment_alm}.
$\gamma$ denotes the largest eigenvalue of the matrix $A^TA$. For the choice of
parameter $\mu$, we take the same strategy as in \cite{YangJ2009-pp} and set
$\mu_0 = 2m / \|\bb\|_1$. We set $\rho=1.5$.

\begin{algorithm}[h]
\caption{\bf (Augmented Lagrange Multiplier Method For Alignment)}
\begin{algorithmic}[1]
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\blamda_1 = \mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $t_{1} \leftarrow 1, z_{1} \leftarrow \x_{k}, \u_{1} \leftarrow \x_{k}$;
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\e_{l+1} \leftarrow \textup{shrink}\left(\bb - A\x_l + \frac{\blamda_k}{\mu_k}, \frac{1}{\mu_k}\right)$;
\STATE $\x_{l+1} \leftarrow (A^\dagger)^T \left(\bb - \e_{l+1} + \frac{\blamda_k}{\mu_k} \right) $;
\ENDWHILE
\STATE $\blamda_{k+1} \leftarrow \blamda_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\STATE $\mu_{k+1} \leftarrow \rho\mu_k$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{algorithmic}
\label{alg:alignment_alm}
\end{algorithm}

\subsection{Computational Complexity} Any discussion of the speed of an
algorithm should discuss the comutational complexity of the algorithm.  For the
alignment stage ALM algorithm \ref{alg:alignment_alm}, this is rather simple.  The
Matrix-vector computions are $O(mn)$, and the vector-vector operations are
$O(m)$, in terms of both computation cost and memory-bandwidth.  Since A is
very tall, the computation of $A^\dagger$, the pseudoinverse of $A$, is best
computed via $A^\dagger = (A^TA)^{-1} A^T$.  The complexity of computing $G =
A^T A$ is $O(nnm)$.  $G$ is sufficiently small that it can be inverted via
Gauss-Jordan elimination, which has a complexity of $O(n^3)$.  The computation
of $A$ finishes with a matrix multiplication, with cost $O(n^2 m)$.
Fortunately, $A^\dagger$ only needs to be computed once per instance of the
$\ell_1$ minimization problem.

Therefore, for problems with large $n$, we expect the computation of the
pseudoinverse to dominate.  However, for the problem sizes needed in face
recognition problems, we find that the computation time is dominated by the
repeated computation of matrix-vector and matrix-matrix multiplications in the
inner loop.

%\STATE Compute residuals $r_i(\bb) = \| {\bb} - {A}_i \, \delta_i(\hat{\x}) \|_2$ for $i = k_1, \dots, k_S$.
